---
title: "RM output to CONSENSUS"
output: rmarkdown::github_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '/Users/rpianezza/TE')
```

Explanation of the RepeatMasker output file:

* `SWscore` = Smith-Waterman score of the match, usually complexity adjusted
* `perc_div` = % substitutions in matching region compared to the consensus
* `perc_del` = % of bases opposite a gap in the query sequence (deleted bp)
* `perc_ins` = % of bases opposite a gap in the repeat consensus (inserted bp)
* `query_sequence` = name of query sequence
* `position_in_query_begin` = starting position of match in query sequence
* `position_in_query_end` = ending position of match in query sequence
* `position_in_query_left` = no. of bases in query sequence past the ending position of match
* `C` = match is with the Complement of the consensus sequence in the database
* `matching_repeat` = name of the matching interspersed repeat
* `repeat_class/family` = the class of the repeat
* `position_in_repeat_begin` = starting position of match in database sequence (using top-strand numbering)
* `position_in_repeat_end` = ending position of match in database sequence
* `position_in_repeat_left` = no. of bases in the repeat consensus sequence 
prior to beginning of the match (so 0 means that the match extended all the way to the end of the repeat consensus sequence)
* `ID` = estimated unique transposon (es. two segments of the same transposon could be separated by another insertion, thus these two sequences have the same ID)
* An asterisk (*) in the final column indicates that there is a higher-scoring match whose domain partly (<80%) includes the domain of this match.

Note that the three column indicating the **position in repeat** are arranged differently for insertions in the two strands:

* For `+` strand: `begin`, `end`, `left`
* For `C` strand: `left`, `end`, `begin`

For simplicity, we kept the order of the `+` strand for all the RM hits and we dealt with this difference in the next analysis calculating `fragment length` and `segment length` differently for the two strands.

Command to remove multiple spaces from the RM output and make it readable in R:
```
less path/rm.out | sed 's/  */ /g' | cut -c2- | > output
```

```{r}
library(tidyverse)
library(ggpubr)
```

```{r}
strains <- c("Es_Ten","Iso1","Pi2","RAL91","RAL176","RAL732","RAL737","SE_Sto")

read_RM_out <- function(path, strain_name) {
  
  out <- read_delim(path, delim = " ", col_names = c("SWscore", "perc_div", "perc_del", "perc_ins", "query_sequence", "position_in_query_begin", "position_in_query_end", "position_in_query_left",  "strand", "matching_repeat", "repeat_class/family", "position_in_repeat_begin", "position_in_repeat_end", "position_in_repeat_left")) %>% select(-X15) %>% mutate(position_in_repeat_begin = str_replace(position_in_repeat_begin, "\\(", "")) %>% mutate(position_in_repeat_begin = str_replace(position_in_repeat_begin, "\\)", "")) %>% mutate(position_in_repeat_left = str_replace(position_in_repeat_left, "\\(", "")) %>% mutate(position_in_repeat_left = str_replace(position_in_repeat_left, "\\)", "")) %>% mutate(position_in_query_left = str_replace(position_in_query_left, "\\(", "")) %>% mutate(position_in_query_left = str_replace(position_in_query_left, "\\)", "")) %>% mutate(position_in_repeat_end = str_replace(position_in_repeat_end, "\\(", "")) %>% mutate(position_in_repeat_end = str_replace(position_in_repeat_end, "\\)", "")) %>% mutate(position_in_query_begin = str_replace(position_in_query_begin, "\\(", "")) %>% mutate(position_in_query_begin = str_replace(position_in_query_begin, "\\)", "")) %>% mutate(position_in_query_end = str_replace(position_in_query_end, "\\(", "")) %>% mutate(position_in_query_end = str_replace(position_in_query_end, "\\)", "")) %>% type_convert()
  
  out_inverted <- out %>% mutate(backup = ifelse(strand == "C", position_in_repeat_begin, "empty"), position_in_repeat_begin = ifelse(strand == "C", position_in_repeat_left, position_in_repeat_begin), position_in_repeat_left = ifelse(strand == "C", backup, position_in_repeat_left)) %>% select(-backup) %>% mutate(strain = strain_name)
  
  out_spoink <- out_inverted %>% filter(matching_repeat == "gypsy-7-sim1")
}

Es_Ten <- read_RM_out("/Volumes/Temp1/simulans-old-strains/Dmel-spoink/consensus/RM-out/readable/D.mel.Es_Ten.fa.ori.out", "Es_Ten")
Iso1 <- read_RM_out("/Volumes/Temp1/simulans-old-strains/Dmel-spoink/consensus/RM-out/readable/D.mel.Iso1.fa.ori.out", "Iso1")
Pi2 <- read_RM_out("/Volumes/Temp1/simulans-old-strains/Dmel-spoink/consensus/RM-out/readable/D.mel.Pi2.fa.ori.out", "Pi2")
RAL91 <- read_RM_out("/Volumes/Temp1/simulans-old-strains/Dmel-spoink/consensus/RM-out/readable/D.mel.RAL91.fa.ori.out", "RAL91")
RAL176 <- read_RM_out("/Volumes/Temp1/simulans-old-strains/Dmel-spoink/consensus/RM-out/readable/D.mel.RAL176.fa.ori.out", "RAL176")
RAL732 <- read_RM_out("/Volumes/Temp1/simulans-old-strains/Dmel-spoink/consensus/RM-out/readable/D.mel.RAL732.fa.ori.out", "RAL732")
RAL737 <- read_RM_out("/Volumes/Temp1/simulans-old-strains/Dmel-spoink/consensus/RM-out/readable/D.mel.RAL737.fa.ori.out", "RAL737")
SE_Sto <- read_RM_out("/Volumes/Temp1/simulans-old-strains/Dmel-spoink/consensus/RM-out/readable/D.mel.SE_Sto.fa.ori.out", "SE_Sto")

RM_merged <- bind_rows(Es_Ten, RAL91, RAL176, RAL732, RAL737, SE_Sto) %>% select(-"matching_repeat", -"repeat_class/family")
```

```{r}
identify_fragments <- RM_merged %>% mutate(length = position_in_query_end-position_in_query_begin,
    distance_from_previous = position_in_query_begin - lag(position_in_query_end, default = first(position_in_query_end)),
    distance_from_next = lead(position_in_query_begin, default = last(position_in_query_begin)) - position_in_query_end,
    HQ_hits = ifelse(SWscore > 10000, "HQ", "fragment"))

fragments_near_fragments <- identify_fragments %>%
  filter(HQ_hits == "fragment" & 
           (lead(HQ_hits) == "fragment" & distance_from_next < 100) | 
           (lag(HQ_hits) == "fragment" & distance_from_previous < 100)) %>% filter(perc_div < 20)
    
fragments_near_HQ <- identify_fragments %>% filter(HQ_hits == "HQ" | (lead(HQ_hits) == "HQ" & distance_from_next < 100) | (lag(HQ_hits) == "HQ" & distance_from_previous < 100)) %>% filter(perc_div < 20) %>% bind_rows(fragments_near_fragments) %>% arrange(strain, query_sequence, position_in_query_begin)

(insertions <- fragments_near_HQ %>% group_by(strain) %>% mutate(insertion_number = cumsum(distance_from_previous > 100) + 1) %>% ungroup() %>% select(-position_in_query_left) %>% group_by(insertion_number, strain) %>% filter(sum(length) > 4500) %>% ungroup() %>% group_by(strain, query_sequence) %>% mutate(insertion_number = cumsum(distance_from_previous > 100) + 1) %>% ungroup() %>% distinct())

insertions_selected <- insertions %>% select(query_sequence, position_in_query_begin, position_in_query_end, strand, position_in_repeat_begin, position_in_repeat_end, strain, distance_from_previous, distance_from_next)

strains <- c("Es_Ten","RAL176","RAL732","RAL737","RAL91","SE_Sto")
for (s in strains){
  single_strain <- insertions_selected %>% filter(strain==s)
  path <- paste0("/Volumes/Temp1/simulans-old-strains/Dmel-spoink/consensus/RM-out/bed/", "D.mel.", s, ".tsv")
  write_tsv(single_strain, path)
}
```

```{}
find /Volumes/Temp1/simulans-old-strains/Dmel-spoink/consensus/RM-out/bed -name '*.tsv' -type f -exec sh -c 'python /Volumes/Temp1/simulans-old-strains/Dmel-spoink/consensus/insertions2bed.py $0 5000 5600 29 ${0%.tsv}.bed' {} \;


for bed_file in /Volumes/Temp1/simulans-old-strains/Dmel-spoink/consensus/RM-out/bed/*.bed; do
    base_name=$(basename "$bed_file" .bed)
    fasta_file="/Volumes/Temp1/simulans-old-strains/Dmel-spoink/consensus/genomes/$base_name.fa"   
    output_fasta="/Volumes/Temp1/simulans-old-strains/Dmel-spoink/consensus/RM-out/fasta/$base_name.fasta"
    bedtools getfasta -s -fi "$fasta_file" -bed "$bed_file" -fo "$output_fasta"
done

cat /Volumes/Temp1/simulans-old-strains/Dmel-spoink/consensus/RM-out/fasta/*.fasta > /Volumes/Temp1/simulans-old-strains/Dmel-spoink/consensus/MSA/spoink-Dmel.fasta

muscle -in /Volumes/Temp1/simulans-old-strains/Dmel-spoink/consensus/MSA/spoink-Dmel.fasta -out /Volumes/Temp1/simulans-old-strains/Dmel-spoink/consensus/MSA/spoink.MSA

python /Volumes/Temp1/simulans-old-strains/Dmel-spoink/consensus/MSA2consensus.py /Volumes/Temp1/simulans-old-strains/Dmel-spoink/consensus/MSA/spoink.MSA /Volumes/Temp1/simulans-old-strains/Dmel-spoink/consensus/MSA/spoink-consensus-dmel.fasta
```
